\documentclass[11pt]{article}
\topmargin=0.0in %length of margin at the top of the page (1 inch added by default)
\oddsidemargin=0.0in %length of margin on sides for odd pages
\evensidemargin=0in %length of margin on sides for even pages
\textwidth=6.5in %How wide you want your text to be
\marginparwidth=0.5in
\headheight=0pt %1in margins at top and bottom (1 inch is added to this value by default)
\headsep=0pt %Increase to increase white space in between headers and the top of the page
\textheight=9.1in %How tall the text body is allowed to be on each page

\usepackage{url}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{wrapfig} % added from executive summary
\usepackage{etaremune}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{listings} 

\begin{document}

\title{CS267 Final Project Proposal}
\author{Frank Austin Nothaft}
\date{}

\maketitle

\section{Proposal}

As seen in homework \#3, genome assembly is a significant challenge. While homework \#3
presented a de Bruijn based genome assembly method~\cite{pevzner01}, ``long read'' assemblers
tend to use overlap-layout concensus~(OLC) based approaches~\cite{rausch09}. As my CS267 project,
I intend to build a distributed OLC assembler using Apache Spark~\cite{zaharia10}. This
assembler targets the PacBio sequencing platform~\cite{eid09} which produces reads of approximately
$\sim10,000$ bases in length.

\section{Components}

This project involves several components:

\begin{enumerate}
\item \textbf{Read Overlapper:} I have already implemented a Spark-based overlapper. This tool uses
a MinHash signature to compute approximate similarity of the reads~\cite{berlin14}.
\item \textbf{Graph Reduction:} Once the reads are overlapped, we need to simplify the overlap graph
of the reads. To do this, I propose to perform a transitive reduction to a string graph-like~\cite{myers05}
structure. I have sketched out how to perform this transitive reduction on a message-passing graph
API~\cite{gonzalez14} using at most two rounds of message passing.
\item \textbf{Assembly Polishing:} Although the PacBio sequencing platform produces very long reads,
the reads have a high error rate. Once we have reduced down to a string graph, we plan to use a statistical
model to pick the best consensus sequence per segment in the graph. We will start with a simple model, such
as a reference-free implementation of the model used in the samtools mpileup variant caller~\cite{li11}.
\end{enumerate}

These components will be implemented on top of the ADAM library~\cite{nothaft15}, which provides primitives for
manipulating genomic datasets inside of Spark.

\section{Evaluation}

We plan to evaluate this project on both performance and accuracy. We will use some of the reference
datasets provided by PacBio~\cite{pb-datasets}, and will run our evaluation on the Amazon EC2 cloud.
We will evaluate:

\begin{itemize}
\item Performance:
\begin{enumerate}
\item Strong scaling on small and large genomes
\item Weak scaling with increased genome size
\item Weak scaling with increased genome coverage
\end{enumerate}
\item Accuracy:
\begin{enumerate}
\item Assembly length (N/NG50)
\item Accuracy against known sequence, where available
\end{enumerate}
\end{itemize}

\bibliographystyle{acm}
\bibliography{cs267_fnothaft_proposal}

\end{document}
