\documentclass[10pt]{article} %Sets the default text size to 11pt and class to article.
%------------------------Dimensions--------------------------------------------
\topmargin=0.0in %length of margin at the top of the page (1 inch added by default)
\oddsidemargin=0.0in %length of margin on sides for odd pages
\evensidemargin=0in %length of margin on sides for even pages
\textwidth=6.5in %How wide you want your text to be
\marginparwidth=0.5in
\headheight=0pt %1in margins at top and bottom (1 inch is added to this value by default)
\headsep=0pt %Increase to increase white space in between headers and the top of the page
\textheight=9.1in %How tall the text body is allowed to be on each page

\usepackage{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{paralist}

\newcounter{pubCtr}

\begin{document}

\centerline {\large\bf Frank Austin Nothaft} 
%\centerline {}
\centerline {\href{mailto:fnothaft@berkeley.edu}{fnothaft@berkeley.edu} $\bullet$ \href{http://www.fnothaft.net}{http://www.fnothaft.net} $\bullet$ 202.340.0466}
\noindent {}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Education}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}
\noindent {{\bf University of California, Berkeley}} \\
\noindent {Doctor of Philosophy, Computer Science. August 2013--present. GPA: 3.81.} \\
\noindent {Advanced to candidacy, May 2016. Advisors: Dave Patterson and Anthony Joseph.} \\
\noindent {Masters of Science, Computer Science. August 2013--May 2015. GPA: 3.79.} \\

\noindent {{\bf Stanford University}} \\
\noindent {Bachelor of Science with Honors, Electrical Engineering. September 2007--June 2011.} \\
\noindent {Minor in Management Science $\&$ Engineering. GPA: 3.24.} \\
\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Honors}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}
\noindent {\bf NSF Graduate Research Fellowship} \\
\noindent {National Science Foundation, August 2013--May 2016} \\

\noindent {\bf Hugh Hildreth Skilling Award for Teaching Excellence} \\
\noindent {Stanford University Department of Electrical Engineering, June 2011} \\

\noindent {\bf Departmental Honors} \\
\noindent {Stanford University Department of Electrical Engineering, June 2011} \\
\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Publications}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}

\noindent {\emph{\textbf{Peer-Reviewed Journal Articles}}}

\setdefaultleftmargin{13pt}{}{}{}{}{}
\begin{enumerate}
\item {John Vivian, Arjun Rao, \textbf{Frank Austin Nothaft}, Christopher Ketchum,
Joel Armstrong, Adam Novak, Jacob Pfeil, Jake Narkizian, Alden D. Deran,
Audrey Musselman-Brown, Hannes Schmidt, Peter Amstutz, Brian Craft, Mary Goldman,
Kate Rosenbloom, Melissa Cline, Brian O'Connor, Megan Hanna, Chet Birger,
W. James Kent, David A. Patterson, Anthony D. Joseph, Jingchun Zhu,
Sasha Zaranek, Gad Getz, David Haussler, and Benedict Paten. ``Toil enables reproducible,
open source, big biomedical data analyses.'' In \emph{Nature
Biotechnology}, April 2017. Originally posted as \emph{BioR$\chi$iv:062497}, July 7, 2016.}
\item Zhao Zhang, Kyle Barbary, \textbf{Frank Austin Nothaft}, Evan Sparks, Oliver Zahn,
Michael J. Franklin, David A. Patterson, and Saul Perlmutter. ``Kira: Processing Astronomy
Imagery Using Big Data Technology.'' In \emph{IEEE Transactions on Big Data},
(IEEE TBD), August 2016. Extended version of Zhang BigData '15.
\item Benedict Paten, Mark Diekhans, Brian J. Druker, Stephen Friend,
Justin Guinney, Nadine Gassner, Mitchell Guttman, W. James Kent, Patrick Mantey,
Adam A. Margolin, Matt Massie, Adam M. Novak, \textbf{Frank Austin Nothaft},
Lior Pachter, David Patterson, Maciej Smuga-Otto, Joshua M. Stuart,
Laura Van’t Veer, Barbara Wold, and David Haussler. ``The NIH BD2K Center for
Big Data in Translational Genomics.'' In \emph{Journal of the American Medical
Informatics Association (JAMIA),} July 2015. Invited.
\setcounter{pubCtr}{\theenumi}
\end{enumerate}

\noindent {\emph{\textbf{Peer-Reviewed Conference Proceedings}}}

\begin{enumerate}
\setcounter{enumi}{\thepubCtr}
\item {Zhao Zhang, Kyle Barbary, \textbf{Frank Austin Nothaft}, Evan Sparks, Oliver Zahn,
Michael J. Franklin, David A. Patterson, and Saul Perlmutter. ``Scientific computing meets big data
technology: An astronomy use case.'' In \emph{Proceedings of the International Conference on
Big Data}, November~2015 (BigData~'15). Originally posted as \emph{Ar$\raisebox{2pt}{$\chi$}$iv:1507.03325.}}
\setcounter{pubCtr}{\theenumi}
\end{enumerate}

\end{minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Publications \\ (con't)}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}

\noindent {\emph{\textbf{Peer-Reviewed Conference Proceedings (con't)}}}

\begin{enumerate}
\setcounter{enumi}{\thepubCtr}
\item {\textbf{Frank Austin Nothaft}, Matt Massie, Timothy Danford, Zhao Zhang,
Uri Laserson, Carl Yeksigian, Jey Kottalam, Arun Ahuja, Jeff Hammerbacher,
Michael Linderman, Michael J. Franklin, Anthony D Joseph, and David A. Patterson.
``Rethinking data-intensive science using scalable analytics systems.'' In
\emph{Proceedings of the International Conference on Management of Data},
May~2015 (SIGMOD~'15).}
\item {\textbf{Frank Austin Nothaft}, Luis Fernandez, Stephen Cefali, Nishant Shah, Luke Darnell,
and Jacob Rael. ``Pragma-based floating-to-fixed point conversion for the emulation of analog
behavioral models.'' In \emph{Proceedings of the International Conference on Computer-Aided Design},
November 2014 (ICCAD '14).}
\item {Krishna Malladi, \textbf{Frank Austin Nothaft}, Kartika Periythambi, Benjamin Lee, 
Christos Kozyrakis, and Mark Horowitz. ``Towards energy-proportional datacenter memory with mobile DRAM.''
In \emph{Proceedings of the International Symposium on Computer Architecture}, June 2012 (ISCA '12).}
\setcounter{pubCtr}{\theenumi}
\end{enumerate}

\noindent {\emph{\textbf{Technical Reports}}}

\begin{enumerate}
\setcounter{enumi}{\thepubCtr}
\item {Matthew Massie, \textbf{Frank Austin Nothaft}, Christopher Hartl, Christos Kozanitis, Andr\'{e}~Schumacher,
Anthony D. Joseph, and David A. Patterson. ``ADAM: Genomics formats and processing patterns
for cloud scale computing.'' \emph{University of California, Berkeley Technical Report
UCB/EECS-2013-207.} November 2013.}
\setcounter{pubCtr}{\theenumi}
\end{enumerate}

\noindent {\emph{\textbf{Theses}}}

\begin{enumerate}
\setcounter{enumi}{\thepubCtr}
\item {\textbf{Frank Austin Nothaft}, ``Scalable genome resequencing with \texttt{ADAM} and \texttt{avocado}.''
Masters Thesis, May 2015. \emph{University of California, Berkeley Technical Report UCB/EECS-2015-65}.} 
\item {\textbf{Frank Austin Nothaft}, ``Design strategies for compiler managed instruction stores.''
Honors Thesis, Stanford University, June 2011.}
\setcounter{pubCtr}{\theenumi}
\end{enumerate}

\noindent {\emph{\textbf{Preprints}}}

\begin{enumerate}
\setcounter{enumi}{\thepubCtr}
\item {Michael D. Linderman, Davin Chia, Forrest Wallace, and \textbf{Frank Austin Nothaft}, ``DECA:
Scalable XHMM exome copy-number variant calling with ADAM and Apache Spark.''
\emph{BioR$\chi$iv:062497}, September 2017.}
\end{enumerate}
  
\noindent{}
\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Presentations}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}

\noindent {\emph{\textbf{Conference Talks}}} \\
\noindent {``Processing 70TB of Genomic Data with ADAM and Toil''} \\
\indent {\emph{Spark Summit}, Boston, MA, February 2017} \\
\indent {\emph{Biological Data Science} (BIODATA '16), Cold Spring Harbor, NY, October 2016} \\
\indent {\emph{Genome Informatics} (GI '16), Hinxton, UK, September 2016} \\
\indent {\emph{Bioinformatics Open Source Conference} (BOSC '16), Orlando, FL, July 2016} \\
\indent {\emph{Spark Summit}, San Francisco, CA, June 2016} \\

\noindent {``Rethinking data-intensive science using scalable analytics systems.''} \\
\indent {\emph{International Conference on Management of Data} (SIGMOD~'15), \\ Melbourne,
Australia, June 2015} \\

\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Presentations (con't)}

\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}

\noindent {\emph{\textbf{Conference Talks (con't)}}} \\
\noindent {``Reproducible Emulation of Analog Behavioral Models.''} \\
\indent {\emph{International Conference on Computer Aided Design} (ICCAD '14), \\ San Jose, CA,
November 2014} \\

\noindent {``ADAM: Fast, Scalable Genome Analysis''} \\
\indent {\emph{Bioinformatics Open Source Conference} (BOSC '14), Boston, MA, July 2014} \\
\indent {\emph{Spark Summit}, San Francisco, CA, June 2014} \\

\noindent {\emph{\textbf{Invited Talks}}} \\
\noindent {``Processing 70TB of Genomic Data with ADAM and Toil''} \\
\indent {\emph{CROSS Research Symposium}, Santa Cruz, CA, October 2016} \\
\indent {\emph{VA Palo Alto}, Palo Alto, CA, October 2016} \\

\noindent {``ADAM: Fast, Scalable Genome Analysis''} \\
\indent {\emph{Color Genomics}, Millbrae, CA, July 2016} \\
\indent {\emph{Johnson and Johnson}, Belgium, December 2015} \\
\indent {\emph{Human Longevity}, Mountain View, CA, June 2015} \\
\indent {\emph{DNANexus}, Mountain View, CA, December 2014} \\
\indent {\emph{Novartis Institutes}, Cambridge, MA, October 2014} \\
\indent {\emph{Wellcome Trust Genome Center}, Hinxton, UK, July 2014} \\

\noindent {``Fast Variant Calling with ADAM and avocado''} \\
\indent {\emph{KTH Kista/SICS}, Kista, Sweden, February 2015} \\

\noindent {``Automation For Validating Behavioral Models Against Schematics'' } \\
\indent {With Nishant Shah, \emph{Cadence Mixed Signal Design Summit}, San Jose, CA,
\\September 2012} \\
\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Work \\ Experience}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}
\centerline {{\bf Databricks, Genomics SME (Consultant) \hfill February 2017--present}}
\noindent {Design and deployment of Apache Spark-based genomics workflows} \\

\centerline {{\bf Broadcom, R\&D Engineer, IC Design 2 \hfill March 2016--June 2016}}
\centerline {{\bf Broadcom, Engineer, Staff 1---IC Design \hfill April 2012--March 2016}}
\centerline {{\bf Broadcom, Engineer---IC Design \hfill June 2011--April 2012}}
\noindent {Design verification and automation for RF/mixed-signal integrated circuits} \\

\noindent {\emph{\textbf{Internships}}} \\
\centerline {\bf NVIDIA, ASIC Intern \hfill Summer 2010}
\noindent {Design validation, test, and characterization for GPU systems} \\

\centerline {\bf SAIC, Systems Engineering Intern \hfill Summer/Winter 2008}
\noindent {Evaluation of technologies for high-reliability emergency telecommunications} \\

\centerline {\bf AJ Engineers, Inc., Electrical Engineering Intern \hfill Summer 2007}
\noindent {Design and drafting of electrical systems for dwellings} \\

\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Teaching}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}

\centerline {{\bf CS162: Operating Systems \hfill Summer 2015, Spring 2016}}
\noindent {Course Assistant for Dr. Charles Reiss, and Professor Anthony D. Joseph, University of California, Berkeley} \\

\centerline {{\bf EE109: Digital Systems Design Lab \hfill Spring 2011}}
\noindent {Course Assistant for Dr. James Weaver, Stanford University} \\

\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Teaching (con't)}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}

\centerline {{\bf EE108A: Digital Systems Design 1 \hfill Fall 2009, 2010, Winter 2010, 2011}}
\noindent {Course Assistant for Professor Subhasish Mitra, Stanford University} \\
\end {minipage}

\begin {minipage}[t]{0.2\linewidth}
\vspace{0pt}
\noindent {\bf Service}
\end {minipage}
\begin {minipage}[t]{0.8\linewidth}
\vspace{0pt}

\noindent {\bf \emph{Standards Bodies}} \\
\noindent {Co-chair, GA4GH Containers and Workflows Working Group, 2015} \\
\noindent {Member, GA4GH Data Working Group, 2014--present} \\
\noindent {}

\noindent {\bf \emph{Conference Organization}}

\noindent {New Frontiers in Computing (NFIC)} \\
\indent {\emph{Co-Chair,} Stanford, CA, 2010} \\
\indent {\emph{Organizing Committee Member,} Stanford, CA, 2009} \\

\noindent {\bf \emph{Reviewing}}

\noindent {Bioinformatics Open Source Conference (BOSC), 2015--2016} \\
\noindent {Hot Topics in Networks Workshop (HotNets), 2014} \\

\noindent {\bf \emph{Professional Society Leadership}}

\noindent {Chair, IEEE Orange County Computer Society, 2013} \\
\noindent {Vice Chair, IEEE Orange County Computer Society, 2012} \\
\noindent {Chair, Stanford University IEEE Student Branch, June 2009--June 2011} \\

\noindent {\bf \emph{Outreach}}

\noindent {Project Mentor} \\
\indent {\emph{Techbridge,} Oakland, CA, 2014} \\

\noindent {Panel on Careers in Science, Technology, Engineering, and Mathematics} \\
\indent {\emph{The Wooden Floor,} Santa Ana, CA, May 2012} \\

\noindent {\bf \emph{Professional Society Membership}}

\noindent {\textbf {IEEE:} Graduate Student Member: 2014--present, Member: 2011--2013,} \\
\noindent {\color{White} \textbf{IEEE:} \color{Black} Student Member: 2007--2011} \\
\noindent {\textbf {ACM:} Member: 2011--present, Student Member: 2011} \\
\noindent {\textbf {ISCB:} Student Member: 2014--present} \\
\noindent {}

\noindent {\bf \emph{Students Mentored}}

\noindent {\emph{Eric Tu,} UC Berkeley, Undergraduate/Masters} \\
\noindent {\emph{Niranjan Kumar,} UC Berkeley, Undergraduate} \\
\noindent {\emph{Ananth Pallaseni,} UC Berkeley, Undergraduate} \\

\end {minipage}

\end{document}
