\documentclass[10pt,oneside]{article}

%%%%%%%%%%%%%
\setlength{\textheight}{8.75in} %Letter is 11in, less 2 for margins, less 0.25 for footer
\setlength{\oddsidemargin}{0.0in} %gets +1inc
\setlength{\evensidemargin}{0.0in} %gets +1inch
\setlength{\textwidth}{6.50in} %Letter is 8.5, less 2 inches for margins
\setlength{\topmargin}{0.0in}
\setlength{\headheight}{0in}
\setlength{\headsep}{0in}
\setlength{\parindent}{0.25in}
%%%%%%%%%%%% 

\usepackage[numbers]{natbib}
\usepackage{graphicx}
\usepackage[colorlinks=true,citecolor=black,urlcolor=blue]{hyperref}

\title{ADAM Enables Distributed Analyses Across \\ Large Scale Genomic Datasets}
\author{\small \underline{Frank~Austin~Nothaft}$^{1,2}$,
Arun~Ahuja$^3$, Timothy~Danford$^{1,4}$, Michael~Heuer$^1$,
Jey~Kottalam$^1$, Matt~Massie$^1$, \and
\small Audrey~Musselman-Brown$^5$, Beau~Norgeot$^{5,6}$, Ravi~Pandya$^7$,
Justin~Paschall$^1$, Jacob~Pfeil$^5$, Hannes~Schmidt$^5$,
\and \small Eric~Tu$^1$, John~Vivian$^5$,
Ryan~Williams$^3$, Carl~Yeksigian$^8$, Michael~Linderman$^3$, Jeff~Hammerbacher$^3$,
\and \small Uri~Laserson$^{3,9}$, Gaddy~Getz$^{10}$, David~Haussler$^5$, Benedict~Paten$^5$, Anthony~D.~Joseph$^1$,
David~A.~Patterson$^{1,2}$}
\date{}

\begin{document}

\footnotetext[1]{AMPLab, University of California, Berkeley, CA}
\footnotetext[2]{ASPIRE Lab, University of California, Berkeley, CA}
\footnotetext[3]{Icahn School of Medicine at Mount Sinai, New York, NY}
\footnotetext[4]{Tamr, Cambridge, MA}
\footnotetext[5]{Genome Informatics Lab, University of California, Santa Cruz, CA}
\footnotetext[6]{University of California, San Francisco, CA}
\footnotetext[7]{Microsoft Research, Redmond, WA}
\footnotetext[8]{GenomeBridge, Cambridge, MA}
\footnotetext[9]{Cloudera, Inc., San Francisco, CA}
\footnotetext[10]{The Broad Institute of MIT and Harvard, Cambridge, MA}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\footnotetext{Correspondence should be addressed to fnothaft@berkeley.edu.}

\maketitle

\vspace{-0.25in}
\noindent
{\small
Website: \url{http://bdgenomics.org} \\
Repositories: \url{https://github.com/bigdatagenomics/adam} and \url{https://github.com/BD2KGenomics/toil-scripts} \\
License: Apache 2 License \\
}

The detection and analysis of rare genomic events requires integrative analysis
across large cohorts with terabytes to petabytes of genomic data. Contemporary
genomic analysis tools have not been designed for this scale of data-intensive
computing. This abstract presents recent updates to ADAM, an Apache 2 licensed
library built on top of the popular Apache Spark distributed computing framework.
We are using ADAM and the Toil workflow management system~(Apache 2 licensed) to
recall the Simons Genome Diversity project dataset against the GRCh38 build of the
human reference genome. Because ADAM is designed to allow genomic analyses to be
seamlessly distributed across large clusters, we achieve a 3.5$\times$ improvement
in end-to-end variant calling latency and a 66\% cost improvement over current toolkits,
without sacrificing accuracy.

On top of our previous results~\cite{nothaft15},
we have achieved an additional 2--$3\times$ improvement in performance by modifying
the schemas that ADAM uses. In addition to improving performance, these modifications
address the problem of storing and processing metadata (e.g., information about a
reference genome build) inline with an analysis. To run our workflow at large scale,
we use the Toil workflow system. Toil is a system for running arbitrary computation that
can be structured as a directed acyclic graph across various different schedulers. Toil
provides fault tolerance, and supports computational reproducibility and portability.
We build upon Toil to enable dynamic scaling of Spark clusters using the AWS spot market.
Our end-to-end variant calling pipeline allocates machines on an as-needed basis to improve
cost effectiveness, and allows the use of Apache Spark in a workflow with traditional, single
node bioinformatics tools.

\bibliographystyle{abbrv}
\bibliography{abstract}

\end{document}
